### 2016-03-24 doc/thesis/intro.tex
Added stuff about model fitting

The branches of the tree may have associated lengths, representing either
evolutionary distance or calendar time between ancestors and their descendants.
The term ``evolutionary distance'' is used here imprecisely to mean any sort of
quantitative measure of evolution, such as the number of differences between
the DNA sequences of an ancestor its descendant, or the difference in average
body mass or height. A phylogeny with branch lengths in calendar time units is
often referred to as \defn{time-scaled}. In a time-scaled phylogeny, the
internal nodes can be mapped onto a timeline by using the tips of the tree,
which usually map to the present day, as a reference
point~\autocite{nee1992tempo}. The corresponding points on the timeline are
called \defn{branching times}, and the rate of their accumulation is referred
to as the \defn{branching rate}. Rooted trees whose tips are all the same
distance from the root are called \defn{ultrametric}
trees~\autocite{buneman1974note}. These concepts are illustrated in
\cref{fig:speciestree}.

@@

In epidemiology, a \defn{transmission tree} is a graphical representation of an
epidemic's progress through a population. Like phylogenies, transmission trees
have tips, nodes, edges, and branch lengths. However, rather than recording an
evolutionary process (speciation), they record an epidemiological process
(transmission). The tips of a transmission tree represent infected hosts, while
internal nodes correspond to transmissions from one host to another.
Transmission trees generally have branch lengths in units of calendar time,
with a root corresponding to the index case, and branching times indicating
times of transmission. The internal nodes may be labelled with
the donor of the transmission pair, if this is known. The tips of the tree,
rather than being fixed at the present day, are placed at the time at which the
individual was removed from the epidemic, such as by death, recovery,
isolation, behaviour change, or migration. Consequently, the transmission tree
may not be ultrametric, but may have tips located at varying distances from the
root. Such trees are said to have \defn{heterochronous}
taxa~\autocite{drummond2003measurably}, in contrast to the
\defn{isochronous} taxa found in the phylogenies of macro-organisms. A
transmission tree is illustrated in \cref{fig:contactnet} (right).

In a transmission tree, each sampled individual appears as a tip when they are
removed from the infected population. In addition, they appear as an internal
node each time they transmit to an individual who is either eventually sampled,
or who transmits directly or indirectly to a sampled individual. This is not
the same as the transmission \emph{network}, in which each infected individual
appears exactly once, and edges are in one-to-one correspondence with
transmissions~\autocite{welch2011statistical, keeling2005networks}.
Transmission networks are discussed further in \cref{sec:contactnet}, and the
distinction between the two objects is illustrated in \cref{fig:contactnet}.
However, since transmission networks generally have no cycles (unless
re-infection occurs), they are trees in the graph theoretical sense, and hence
are sometimes also referred to as transmission
trees~\autocite[\eg][]{kenah2015algorithms}. In this work, we reserve the term
``transmission tree'' for the objects depicted on the right side of
\cref{fig:contactnet}, following~\autocite[\eg][]{stadler2013uncovering}. The
term ``transmission network'' is taken to mean the subgraph of the contact
network along which transmissions occured,
following~\autocite{welch2011statistical, keeling2005networks}.

@@

In general, \defn{viral phylogenies} are simply phylogenetic trees relating
virus strains. In phylodynamics, we often consider \defn{inter-host}
phylogenies, which relate one viral genotype from each host in a population.
The tips of the tree are labelled with both the virus and its host. The crux of
\defn{phylodynamics}~\autocite{grenfell2004unifying} is the fact that the
processes generating these two types of tree - epidemiological processes, for
transmission trees, and evolutionary processes, for viral phylogenies - occur
on similar time scales for RNA viruses~\autocite{drummond2003measurably}. As a
result, there is a close relationship between the two trees. In particular, the
transmission process is quite similar to \defn{allopatric
speciation}~\autocite{coyne2004speciation}, where genetic divergence follows
the geographic isolation of a sub-population of organisms. Thus, transmission,
which is represented as branching in the transmission tree, causes branching in
the viral phylogeny as well. Similarly, the removal of an individual causes the
extinction of their viral lineage. Due to these relationships, the topology of
the viral phylogeny is often used as a proxy for the topology of the
transmission tree. However, there are several complications and caveats which
must be kept in mind when estimating the transmission tree in this manner.

First are the issues of rooting and time-scaling. Modern likelihood-based
methods of phylogenetic reconstruction produce unrooted trees whose branch
lengths measure genetic distance in units of expected substitutions per site,
whereas transmission trees are rooted with branches measuring calendar time. We
generally assume that sampling a virus from the individual also corresponds to
their removal from the transmission tree, so the positions of the tips in time
are fixed. Therefore, to transform a viral phylogeny into an estimated
transmission tree, we must find a root and an assignment of branch lengths such
that the tips are placed at their proscribed times. Of course, there are an
infinite number of possible combinations of root and branch lengths which
result in the correct placement of the tips, so we would also like to find
values which are the most ``reasonable'' given our data. By reasonable, we mean
that the variation in \defn{evolutionary rate}, which is the ratio of a
branch's length in genetic distance to its length in calendar time, is low
across the tree. While there is some variation among hosts, due to
immunological and other factors, we generally expect to observe globally
similar evolutionary rates. Methods for time-scaling a phylogeny include
root-to-tip regression~\autocite{shankarappa1999consistent, korber2000timing,
drummond2003inference}, which we apply in this work, and least-square
dating~\autocite{to2015fast}. Both of these methods can be used to root the
tree, by simply trying all possible root positions and choosing the one which
minimizes a loss function ($1-R^2$, or root-mean-square error, for root-to-tip
regression; sum of squared errors for least-square dating). Alternatively, the
tree may be rooted with an outgroup~\autocite{li1988rates} before time-scaling.

@@

What we have just discussed is a two-step procedure for estimating the
transmission tree. First, a viral phylogeny is constructed from genetic
sequence data, and then it is rooted and time-scaled into a transmission tree.
This approach is straightforward, frequently used, and has the advantage of
leveraging tried-and-true tools for phylogenetic inference. However, it also
has drawbacks, perhaps the most obvious being the multiplication of errors
produced by the separate steps. One commonly used alternative method is to
directly estimate a time-scaled phylogeny by simultaneously inferring the tree
topology, its root and branch lengths, and the parameters of a \defn{molecular
clock} model. A molecular clock is a hypothesis about the evolutionary rates
along the branches of the tree, such as that they are all equal (a \defn{strict
clock}) or that they are \gls{iid} from a common distribution (a \defn{relaxed
clock}). This inference is usually done in a Bayesian framework using
\gls{MCMC}, so that prior information (including the tip dates) 
%\comment{are tip dates part of the prior?} 
can be included in the analysis, and the so-called nuisance parameters of the
molecular clock model can be marginalized out. Software packages for performing
these analyses include BEAST~\autocite{bouckaert2014beast} and
MrBayes~\autocite{ronquist2012mrbayes}.

@@

The contact network underlying an epidemic constrains the shape of the
transmission network, which in turn determines the topology of the transmission
tree relating the infected hosts (\cref{fig:contactnet}). The initially
infected node who introduces the epidemic into the network, also known as the
\defn{index case}, becomes the root of the tree. Each time a transmission
occurs, the lineage corresponding to the donor host in the tree splits into
two, representing the recipient lineage and the continuation of the donor
lineage. \Cref{fig:contactnet} illustrates this correspondence. It's important
to note that, although the order and timing of transmissions determines the
tree topology uniquely, the converse does not hold. That is, for any given
topology, there are in general many transmission networks which would lead to
that topology. In other words, it impossible to distinguish who transmitted to
whom from a transmission tree alone~\autocite{bernard2007hiv}.

@@

\section{Approximate Bayesian computation}
\label{sec:abc}

\subsection{Overview and motivation}

\Gls{ABC} is a statistical method designed to fit complex models, which cannot
be fit using more conventional methods, to observed
data~\autocite{marin2012approximate, sunnaker2013approximate,
beaumont2010approximate} We shall make this precise below, but to fully
describe and motivate \gls{ABC}, it is necessary to first explain exactly what
we mean by ``fitting'' a model. We will illustrate this concept with one of the
simplest and most ubiquitous models: linear regression.

A linear regression is a model of the relationship between some data $\vec{x} =
(x_1, \ldots, x_n)$ and outcomes $\vec{y} = (y_1, \ldots, y_n)$. This model
assumes that the outcomes are linearly related to the data, modulo some noise
introduced by, say, measurement errors and environmental fluctuations. It other
words, there is a constant $\beta$ such that $y_i = \beta x_i + \varepsilon_i$,
where $\varepsilon_i$ is the error associated with the outcome $y_i$. If we
make the additional assumption that the errors are normally distributed, then
the model takes the form
\[
  y_i = \beta x_i + \N(0, \sigma^2),
\]
where $\N(0, \sigma^2)$ indicates a normally distributed random variable with
mean 0 and variance $\sigma$. We will denote the \gls{pdf} of this random
variable by $f_\N(\cdot \mid \mu, \sigma)$. In this formulation, the
coefficient $\beta$ and the error variance $\sigma$ are the two parameters of
the model. If the $y_i$ are all independent, then for particular choice of
$\beta$ and $\sigma$, we can write down the probability density of observing
$\vec{y}$ given $\vec{x}$.
\[
  f(\vec{y} \mid \vec{x}, \beta, \sigma) = 
  \prod_{i=1}^n f_\N(y_i - \beta x_i \mid 0, \sigma^2).
\]
The higher the value of this probability density, the more likely the
observations $\vec{y}$ are given $\vec{x}$ under the model. This gives us a
natural criterion for choosing the parameters: we want to pick $\beta$ and
$\sigma$ which define a model where the probability of our observed data is as
high as possible. When performing such an optimization for fixed data, the
density function just written is also called the \defn{likelihood} of the
parameters,
\[
  \L(\beta, \sigma \mid \vec{x}, \vec{y}) = f(\vec{y} \mid \vec{x}, \beta, \sigma).
\]
The particular $\beta$ and $\sigma$ which optimize $\L(\beta, \sigma \mid
\vec{x}, \vec{y})$ are called the \gls{ML} estimates.

\Gls{ML} inference makes use only of the data at hand, in this case
$\vec{x}$ and $\vec{y}$, to estimate the parameters $\beta$ and $\sigma$.
However, it is frequently the case that the investigator has some additional
\defn{prior} information or belief about what $\beta$ and $\sigma$ are likely
to be. For example, the instrument used to measure the $\vec{y}$ may have a
known margin of error, or the sign of the slope may be obvious from previous
experiments. A Bayesian approach to fitting this model would make use of this
information by codifying the investigator's beliefs as a \defn{prior
distribution}, denoted $f(\beta, \sigma)$. The optimal parameters are then
those which maximize the product of the prior and the likelihood,
\[
  f(\vec{y} \mid \vec{x}, \beta, \sigma) f(\beta, \sigma).
\]
Bayes' theorem tells us that this product is proportional to the
\defn{posterior distribution} $f(\beta, \sigma \mid \vec{y}, \vec{x})$. The
parameters which optimize the posterior, and hence also optimize the above
product, are called the \gls{MAP} estimates.

Now that we have defined what it means to fit a linear regression, there
remains the question of how to go about finding this optimal $\beta$ value.
Na\"{i}vely, we could simply try many different $\beta$ and $\sigma$ values,
calculate the likelihood of each, and choose those which yield the highest
value. This approach, though basic, falls under the umbrella of \defn{numerical
optimization} of the likelihood function. Of course, there is a whole field
devoted to more sophisticated methods for exploring the parameter space, but
they all boil down to the basic idea of trying out different values and
choosing those which give the highest likelihood.

In the case of linear regression, there is another method we could use. We know
from calculus that the maximum of $\L$ occurs either on the boundary of its
domain, or at a point where its partial derivatives with respect to $\beta$ and
$\sigma$ are both zero. Though we will not show it here, it turns out to be
straightforward to find the $\beta$ and $\sigma$ values where the partial
derivatives are zero. In fact, these values co-incide with the least-squares
estimates of $\beta$ and $\sigma$, which minimize the sum of the squared $x_i -
\beta y_i$. However, the method of directly minimizing the loss function with
calculus is only applicable to a narrow class of simple models. For the
majority, the likelihood function is too complicated to set the partial
derivatives with respect to the parameters to zero and solve.

An approximate Bayesian computation approach to least-squares regression might
be as follows. We will need to make an additional assumption: that the errors
$\varepsilon_i$ are normally distributed. That is, our linear model is now of
the form
\[
  y_i = bx_i + \N(0, \sigma),
\]
where $\N(0, \sigma)$ indicates a normal distribution with mean zero and
variance $\sigma$. We do not know $\sigma$ in advance, so this is another
parameter we will need to estimate. There is nothing special about normal
distributions - we could have chosen any other distribution we deemed
appropriate.

The main idea of \gls{ABC} is to replace the traditional likelihood-based
criteria for model ``goodness'' with one based on simulated data generated by
the model. Good models should be able to simulate data which closely resemble
reality. In \gls{ABC}, we are no longer interested in the posterior
distribution, but rather in the distribution of model parameters which produce
data sufficiently ``close'' to the real data. To formalize this, let $d(\cdot,
\cdot)$ be a distance function on datasets, and $\varepsilon$ be a tolerance
level. The objective of \gls{ABC} is to recover the distribution\ldots

@@

Approximate Bayesian computation does not refer to a particular procedure for
model fitting. Rather, ABC refers to the general strategy of choosing model
parameters based on the resulting model's propensity to generate data
resembling the real data. There are three main classes of ABC algorithm which
have been developed so far: rejection, \gls{MCMC}, and sequential Monte Carlo
(SMC).

>>>>

The branches of the tree may have associated lengths, representing either
evolutionary distance or calendar time between ancestors and their descendants.
The term ``evolutionary distance'' is used here imprecisely to mean any sort of
quantitative measure of evolution, such as the number of differences between
the DNA sequences of an ancestor its descendant, or the difference in average
body mass or height. A phylogeny with branch lengths in calendar time units is
often referred to as \defn{time-scaled}. In a time-scaled phylogeny, the
internal nodes can be mapped onto a timeline by using the tips of the tree,
which usually correspond to the present day, as a reference
point~\autocite{nee1992tempo}. The corresponding points on the timeline are
called \defn{branching times}, and the rate of their accumulation is referred
to as the \defn{branching rate}. Rooted trees whose tips are all the same
distance from the root are called \defn{ultrametric}
trees~\autocite{buneman1974note}. These concepts are illustrated in
\cref{fig:speciestree}.

@@

In epidemiology, a \defn{transmission tree} is a graphical representation of an
epidemic's progress through a population. Like phylogenies, transmission trees
have tips, nodes, edges, and branch lengths. However, rather than recording an
evolutionary process (speciation), they record an epidemiological process
(transmission). The tips of a transmission tree represent infected hosts, while
internal nodes correspond to transmissions from one host to another.
Transmission trees generally have branch lengths in units of calendar time,
with branching times indicating times of transmission. The root of a
transmission tree corresponds to the initially infected patient who introduced
the epidemic into the network, also known as the \defn{index case}. The
internal nodes may be labelled with the donor of the transmission pair, if this
is known. The tips of the tree, rather than being fixed at the present day, are
placed at the time at which the individual was removed from the epidemic, such
as by death, recovery, isolation, behaviour change, or migration. Consequently,
the transmission tree may not be ultrametric, but may have tips located at
varying distances from the root. Such trees are said to have
\defn{heterochronous} taxa~\autocite{drummond2003measurably}, in contrast to
the \defn{isochronous} taxa found in most phylogenies of macro-organisms. A
transmission tree is illustrated in \cref{fig:contactnet} (right).

Due to the internal nodes, each infected individual in an epidemic may appear
in the transmission tree more than once. This is different from the
transmission \emph{network}, in which each infected individual appears exactly
once, and edges are in one-to-one correspondence with
transmissions~\autocite{welch2011statistical, keeling2005networks}.
Transmission networks are discussed further in \cref{sec:contactnet}, and the
distinction between the two objects is illustrated in \cref{fig:contactnet}.
However, since transmission networks generally have no cycles (unless
re-infection occurs), they are trees in the graph theoretical sense, and hence
are sometimes also referred to as transmission
trees~\autocite[\eg][]{kenah2015algorithms}. In this work, we reserve the term
``transmission tree'' for the objects depicted on the right side of
\cref{fig:contactnet}, following~\autocite[\eg][]{stadler2013uncovering}. The
term ``transmission network'' is taken to mean the subgraph of the contact
network along which transmissions occured,
following~\autocite{welch2011statistical, keeling2005networks}.

@@

In general, \defn{viral phylogenies} are simply phylogenetic trees relating
virus strains. In phylodynamics, we often consider \defn{inter-host}
phylogenies, which relate one viral genotype from each host in a population.
The crux of \defn{phylodynamics}~\autocite{grenfell2004unifying} is the fact
that the epidemiological processes recorded in transmission trees, and the
evolutionary processes recorded in viral phylogenies, occur on similar time
scales for RNA viruses~\autocite{drummond2003measurably}. As a result, there is
a close relationship between the two types of tree. In particular, the
transmission process is quite similar to \defn{allopatric
speciation}~\autocite{coyne2004speciation}, where genetic divergence follows
the geographic isolation of a sub-population of organisms. Thus, transmission,
which is represented as branching in the transmission tree, causes branching in
the viral phylogeny as well. Similarly, the removal of an individual from the
transmission tree causes the extinction of their viral lineage in the
phylogeny. Due to these relationships, the topology of the viral phylogeny is
often used as a proxy for the topology of the transmission tree. However, there
are several complications and caveats which must be kept in mind when
estimating the transmission tree in this manner.

First are the issues of rooting and time-scaling. Modern likelihood-based
methods of phylogenetic reconstruction~\autocite[\eg][]{price2010fasttree,
stamatakis2014raxml} produce unrooted trees whose branch lengths measure
genetic distance in units of expected substitutions per site. On the other
hand, transmission trees are rooted, and have branches measuring calendar
time~\autocite{pybus2009evolutionary}. It is generally assumed that sampling a
virus from the individual also corresponds to their removal from the
transmission tree, so the positions of the tips in time are fixed. Therefore,
to transform a viral phylogeny into an estimated transmission tree, we must
find a root and an assignment of branch lengths such that the tips are placed
at their proscribed times. In addition, the branch lengths should be chosen
such that the variation in \defn{evolutionary rate}, which is the ratio of a
branch's length in genetic distance to its length in calendar time, is low
across the tree. While there is some variation among hosts, due to
immunological and other factors, we generally expect to observe globally
similar evolutionary rates. Methods for time-scaling a phylogeny include
root-to-tip regression~\autocite{shankarappa1999consistent, korber2000timing,
drummond2003inference}, which we apply in this work, and least-square
dating~\autocite{to2015fast}. Both of these methods can be used to root the
tree, by simply trying all possible root positions and choosing the one which
minimizes a loss function ($1-R^2$, or root-mean-square error, for root-to-tip
regression; sum of squared errors for least-square dating). Alternatively, the
tree may be rooted separately with an outgroup~\autocite{li1988rates} before
time-scaling.

@@

What we have just discussed is a two-step procedure for estimating the
transmission tree. First, a viral phylogeny is constructed from genetic
sequence data, and then it is rooted and time-scaled into a transmission tree.
This approach is straightforward, frequently used, and has the advantage of
leveraging tried-and-true tools for phylogenetic inference. However, it also
has drawbacks, perhaps the most obvious being the multiplication of errors
produced by the separate steps. One commonly used alternative method is to
directly estimate a time-scaled phylogeny by simultaneously inferring the tree
topology, its root and branch lengths, and the parameters of a \defn{molecular
clock} model. A molecular clock is a hypothesis about the evolutionary rates
along the branches of the tree, such as that they are all equal (a \defn{strict
clock}) or that they are \gls{iid} from a common distribution (a \defn{relaxed
clock}). This inference is usually done in a Bayesian framework using
\gls{MCMC}, so that prior information (including the tip dates) can be included
in the analysis, and the so-called nuisance parameters of the molecular clock
model can be marginalized out. Software packages for performing these analyses
include BEAST~\autocite{bouckaert2014beast} and
MrBayes~\autocite{ronquist2012mrbayes}.

@@

The contact network underlying an epidemic constrains the shape of the
transmission network, which in turn determines the topology of the transmission
tree relating the infected hosts (\cref{fig:contactnet}). The index case who
introduces the epidemic into the network becomes the root of the tree. Each
time a transmission occurs, the lineage corresponding to the donor host in the
tree splits into two, representing the recipient lineage and the continuation
of the donor lineage. \Cref{fig:contactnet} illustrates this correspondence.
It's important to note that, although the order and timing of transmissions
determines the tree topology uniquely, the converse does not hold. That is, for
any given topology, there are in general many transmission networks which would
lead to that topology. In other words, it impossible to distinguish who
transmitted to whom from a transmission tree alone~\autocite{bernard2007hiv}.

@@

\section{Approximate Bayesian computation}
\label{sec:abc}
\glsreset{ABC}

\subsection{Model fitting}

A \defn{mathematical model} is a formal description of a hypothesized
relationship between some observed data, $\vec{x} = \set{x_1, \ldots, x_n}$,
and outcomes, $\vec{y} = \set{y_1, \ldots, y_n}$. A \defn{parametric} model
defines a family of possible relationships between data and outcomes, indexed
by one or more numeric parameters $\theta$. A \defn{statistical} model
describes the relationship between data and outcomes in terms of probabilities.
Statistical models define, either explicitly or implicitly, the probability of
observing $\vec{y}$ given $\vec{x}$ and, if the model is parameteric, $\theta$.
In this context, the observed outcomes are taken to be realizations of random
variables $\vec{Y} = \set{Y_1, \ldots, Y_n}$. Note that it is entirely possible
to have no data $\vec{x}$, only observed outcomes $\vec{y}$. In this case, a
model would describe the process by which $\vec{y}$ is generated.

To illustrate these concepts, consider the well-known linear model. For
clarity, we will restrict our attention to the case of one-dimensional data and
outcomes where each $x_i$ and $y_i$ is a real number. The linear model
postulates that the outcomes are linearly related to the data, modulo some
noise introduced by measurement error, environmental fluctuations, and other
external factors. Formally, $y_i = \beta x_i + \varepsilon_i$, where $\beta$ is
the slope of the linear relationship, and $\varepsilon_i$ is the error
associated with measurement $i$. We can make this model a statistical one by
hypothesizing a distribution for the error terms $\varepsilon_i$; most
commonly, it is assumed that they are normally distributed with variance
$\sigma$. In mathematical terms, $Y_i \sim \beta x_i + \N(0, \sigma^2)$, where
``$\sim$'' means ``is distributed as''. We can see from this formulation that
the model is parametric, with parameters $\beta$ and $\sigma$. Moreover, we can
write down the probability density of observing outcome $y_i$ given the
parameters,
\[
  \Pr(Y_i = y_i \mid \beta, \sigma) = 
  \Pr\left(\N(0, \sigma^2) = y_i - \beta x_i\right).
\]
Note that we have followed the standard statistical abuse of notation and used
``$\Pr$'' to refer to a probability \emph{density}, rather than a probability.
Assuming all the $y_i$ are independent, the probability density of the entire
observed set of outcomes is the product of the probability density of each
individual $y_i$,
\[
  \Pr(\vec{Y} = \vec{y} \mid \beta, \sigma) = 
  \prod_{i=1}^N \Pr(Y_i = y_i \mid \beta, \sigma).
\]

For a general model, the probability density of $\vec{y}$ given the parameters
$\theta$ is also known as the \defn{likelihood}, written $\L$, of $\theta$.
That is, $\L(\theta \mid \vec{y}) = \Pr(\vec{y} \mid \theta)$. The higher the
value of the likelihood, the more likely the observations $\vec{y}$ are under
the model. Thus, the likelihood provides a natural criterion for fitting the
model parameters: we want to pick $\theta$ such that the probability density of
our observed outcomes $\vec{y}$ is as high as possible. The parameters which
optimize the likelihood are known as the \textit{\gls{ML}} estimates, denoted
$\hat{\theta}$. \Gls{ML} estimation is usually performed with numerical
optimization. In the simplest terms, many possible values for $\theta$ are
examined, $\L(\theta \mid \vec{y})$ is calculated for each, and the parameters
which produce the highest value are accepted. Many sophisticated numerical
optimization methods exist, although they may not be guaranteed to find the
true \gls{ML} estimates if the likelihood function is complex. Occasionally, as
in the case of least squares, the \gls{ML} estimates can be found explicitly by
setting the likelihood function's derivatives to zero.

\Gls{ML} estimation makes use only of the data and outcomes to estimate the
model parameters $\theta$. However, it is frequently the case that the
investigator has some additional information or \defn{prior belief} about what
$\theta$ are likely to be. For example, in the linear regression case, the
intstrument used to measure the outcomes may have a well-known margin of error,
or the sign of the slope may be obvious from previous experiments. The Bayesian
approach to model fitting makes use of this information by codifying the
investigator's beliefs as a \defn{prior distribution} on the parameters,
denoted $\Pr(\theta)$. Instead of considering only the likeilhood, Bayesian
inference focuses on the product of the likelihood and the prior, $\Pr(\vec{y}
\mid \theta) \Pr(\theta)$. Bayes' theorem tells us that this product is related
to the \textit{posterior distribution} on $\theta$,
\[
  \Pr(\theta \mid \vec{y}) = \frac{\Pr(\vec{y} \mid \theta) \Pr(\theta)}
                                  {\Pr(\vec{y})}.
\]

In principle, $\Pr(\vec{y} \mid \theta) \Pr(\theta)$ can be optimized
numerically just like $\L(\theta \mid \vec{y})$, which would also optimize the
posterior distribution. The resulting optimal parameters are called the
\gls{MAP} estimates. However, from a Bayesian perspective, $\theta$ is not a
fixed quantity to be estimated, but rather a random variable with an associated
distribution (the posterior). Therefore, the \gls{MAP} estimate by itself is of
limited value without associated statistics about the posterior distribution,
such as the mean or credible intervals. Unfortunately, to calculate such
statistics, it is necessary to evaluate the normalizing constant
$\Pr(\vec{y})$, which is almost always an intractible integral.

A popular method for circumventing the normalizing constant is the use of
\gls{MCMC} to obtain a sample from the posterior distribution. \Gls{MCMC} works
by defining a Markov chain whose states are indexed by possible model
parameters. The transition probability from state $\theta_1$ to state
$\theta_2$ is taken to be
\[
  \max\left(1, \frac{\Pr(\vec{y} \mid \theta_2) \Pr(\theta_2) q(\theta_2, \theta_1)}
                    {\Pr(\vec{y} \mid \theta_1) \Pr(\theta_2) q(\theta_1, \theta_2)} \right),
\]
where $q(\theta, \theta')$ is a symmetric \defn{proposal distribution} used in
the algorithm to generate the chain. The stationary distribution of this Markov
chain is equal to the posterior distribution on $\theta$. Therefore, if a long
enough random walk is performed on the chain, the distribution of states
visited will be a Monte Carlo approximation of $\Pr(\theta \mid \vec{y})$, from
which we can calculate statistics of interest. Actually performing this random
walk is straightforward and is accomplished via the Metropolis-Hastings
algorithm.

\begin{algorithm}
  \begin{algorithmic}
    \State Draw $\theta$ according to $\Pr(\theta)$
    \Loop
      \State Propose $\theta'$ according to $q(\theta, \theta')$
      \State Accept $\theta \gets \theta'$ with probability
      $\max \left( 1, 
       \dfrac{\Pr(\vec{y} \mid \theta') \Pr(\theta') q(\theta', \theta)}
             {\Pr(\vec{y} \mid \theta\phantom{'}) \Pr(\theta\phantom{'}) q(\theta, \theta')}
       \right)$
    \EndLoop
  \end{algorithmic}
  \label{alg:mh}
\end{algorithm}

\subsection{Overview and motivation for ABC}

Most mathematical models are amenable to fitting via one or both of the
approaches, \gls{ML} or Bayesian inference, discussed above. However, there are
some, particularly in the domain of population genetics, for which calculation
of either the likelihood or the product of the likelihood and the prior may be
infeasible. For example, one or both of these quantities may be expressible
only as an intractible integral. \Gls{ABC} is designed for such cases, where
standard likelihood-based techniques for model fitting cannot be applied.

The first practical application of \gls{ABC} was by
\textcite{tavare1997inferring}, who wanted to calculate the posterior
distribution on the coalescence time $t$ of $n$ sequences with $k$ observed
segregating sites under the coalescent model~\autocite{kingman1982coalescent}.
That is,
\[
  \Pr(t \mid k) \propto \Pr(t) \Pr(k \mid t),
\]
where again we abuse $\Pr$ to refer to a probability density.
\citeauthor{tavare1997inferring} observed that the expected number of
segregating sites depends on the total branch length of the tree, rather than
its height. This enabled them to write an alternative expression for the
product of the likelihood and prior in terms of the total branch length $l$,
\[
  \int_0^\infty \Pr(t, l) \Pr(k \mid t, l) \d l.
\]
While it is unclear how to evaluate this integral, the power of this
formulation is the fact that it is straightforward to \emph{simulate} points
according to $\Pr(t, l)$, the joint probability of $t$ and $l$, by simply
running the (stochastic) coalescent process and observing the values $\bar{t}$
and $\bar{l}$ which result.

%\Gls{ABC} is a statistical method designed to fit complex models, which cannot
%be fit using more conventional methods, to observed
%data~\autocite{marin2012approximate, sunnaker2013approximate,
%beaumont2010approximate}. We shall make this precise below, but to fully
%describe and motivate \gls{ABC}, it is necessary to first explain exactly what
%we mean by ``fitting'' a model. We will illustrate this concept with one of the
%simplest and most ubiquitous models: linear regression.
%
%An approximate Bayesian computation approach to least-squares regression might
%be as follows. We will need to make an additional assumption: that the errors
%$\varepsilon_i$ are normally distributed. That is, our linear model is now of
%the form
%\[
%  y_i = bx_i + \N(0, \sigma),
%\]
%where $\N(0, \sigma)$ indicates a normal distribution with mean zero and
%variance $\sigma$. We do not know $\sigma$ in advance, so this is another
%parameter we will need to estimate. There is nothing special about normal
%distributions - we could have chosen any other distribution we deemed
%appropriate. 
%
%The main idea of \gls{ABC} is to replace the traditional likelihood-based
%criteria for model ``goodness'' with one based on simulated data generated by
%the model. Good models should be able to simulate data which closely resemble
%reality. In \gls{ABC}, we are no longer interested in the posterior
%distribution, but rather in the distribution of model parameters which produce
%data sufficiently ``close'' to the real data. To formalize this, let $d(\cdot,
%\cdot)$ be a distance function on datasets, and $\varepsilon$ be a tolerance
%level. The objective of \gls{ABC} is to recover the distribution\ldots

@@

\Gls{ABC} does not refer to a particular procedure for model fitting. Rather,
\gls{ABC} refers to the general strategy of choosing model parameters based on
the resulting model's propensity to generate data resembling the real data.
There are three main classes of \gls{ABC} algorithm which have been developed
so far: rejection, \gls{MCMC}, and \gls{SMC}.
